---
title: "DS4002 Pokemon Project Data Cleaning"
author: "Caroline Clippinger"
date: "`r Sys.Date()`"
output: github_document
---

# General Information
This data cleaning script is purposed with preparing the Pokemon type x word and Pokemon generation x word matrices to use for our hierarchical clustering analysis. The necessary data file can be found at DS-4002-group-01/project1/DATA/pokedex.csv in the github. It also requires the following packages: `here`, `tidyverse`, `tidytext`, `textdata`, and `udpipe`. To install these packages use the `install.packages("package_name")` syntax.


# Loading Packages and Data
```{r setup}
knitr::opts_chunk$set(echo = TRUE)
library(here) # for loading data
library(tidyverse) # for general data wrangling and string manipulation
library(tidytext) # for working with the text
library(textdata) # for afinn lexicon to do sentiment analysis
library(udpipe) # for the smart word cleaning and lemmatization


pokemon <- read.csv(here("project1/DATA/", "pokedex.csv"))
```

# Data Cleaning
## Adressing Missing Type Data
```{r}
# Checking for pokemon missing type information
cat("The following", length(pokemon[pokemon$type_1 == "" | is.na(pokemon$type_1), "name"]), "pokemon have missing type information:\n", paste(pokemon[pokemon$type_1 == "" | is.na(pokemon$type_1), "name"]))

# Researching the type information for those pokemon and creating a reference table
researched_types <- tribble(
  ~name, ~type_1, ~type_2,
  "deoxys", "psychic", NA,
  "wormadam", "bug", NA, # depending on cloak second type changes so leaving it empty since its basic cloak is bug
  "giratina", "ghost", "dragon", 
  "shaymin", "grass", NA, # using the sky form type because description talks about flowering plant
  "basculin", "water", NA,
  "darmanitan", "fire", NA,
  "frillish", "water", "ghost", 
  "jellicent", "water", "ghost",
  "tornadus", "flying", NA,
  "thundurus", "electric", "flying",
  "landorus", "ground", "flying",
  "keldeo", "water", "fighting", 
  "meloetta", "normal", "psychic", # this is default form types
  "pyroar", "fire", "normal",
  "meowstic", "psychic", NA,
  "aegislash", "steel", "ghost",
  "pumpkaboo", "ghost", "grass",
  "gourgeist", "ghost", "grass",
  "zygarde", "dragon", "ground",
  "oricorio", "fire", "flying", # based on the color and description i can assume its referencing the fire baile style form
  "lycanroc", "rock", NA,
  "wishiwashi", "water", NA,
  "minior", "rock", "flying",
  "mimikyu", "ghost", "fairy", 
  "toxtricity", "electric", "poison",
  "eiscue", "ice", NA,
  "indeedee", "psychic", "normal",
  "morpeko", "electric", "dark", 
  "urshifu", "fighting", "dark", # looked up description depending on form to determine if it is dark or water
  "basculegion", "water", "ghost", 
  "enamorus", "fairy", "flying",
  "oinkologne", "normal", NA,
  "maushold", "normal", NA,
  "squawkabilly", "normal", "fighting",
  "palafin", "water", NA,
  "tatsugiri", "dragon", "water",
  "dudunsparce", "normal", NA
)

# Updating types with reference table and checking if still no missings
pokemon_updated <- pokemon %>% 
  rows_update(researched_types, by = "name", unmatched = "ignore")

cat("\n\nAfter updating the data with the reference table there are now these pokemon missing type information:\n")
print(pokemon_updated[pokemon_updated$type_1 == "" | is.na(pokemon_updated$type_1), "name"])
```

## Extracting Words Using NLP
```{r}
# Loading English NLP model for tokenizing data
ud_model <- udpipe_download_model(language = "english")

# Extracting the file model
ud_model <- udpipe_load_model(ud_model$file_model)

# Conducting preliminary pokedex entry cleaning to prevent confusing the model
pokemon_smart_clean <- pokemon_updated %>% 
  mutate(
    pokedex_entry_clean = str_replace_all(pokedex_entry, "é", "e"),
    pokedex_entry_clean = str_remove_all(pokedex_entry_clean, "[“”'’‘\"−°—-]"),
    pokedex_entry_clean = str_replace_all(pokedex_entry, "an other", "another"), # noticed this typo when exploring entries
    pokedex_entry_clean = str_replace_all(pokedex_entry, "elec tricity", "electricity") # noticed this typo when exploring entries
  )

# "Annotating" the pokedex entry data, dividing documents by the pokemon name
entry_analysis <- udpipe_annotate(
  ud_model, # model loaded then extracted earlier
  x = pokemon_smart_clean$pokedex_entry_clean, # the text we want the NLP to run on 
  doc_id = pokemon_smart_clean$name, # how to differentiate documents (in our case pokedex entries)
) %>% 
  as.data.frame() # converting to data frame

# Selecting relevant columns from the NLP annotation and applying filtering
pokemon_words <- entry_analysis %>% 
  select(name = doc_id, entry_sentence = sentence, word_og = token, word_root = lemma, pos = upos) %>% 
  filter(!pos %in% c("PUNCT","NUM")) %>% # We do not care about punctuation or numbers as 'descriptive language'
  filter(!str_detect(word_root, "'s")) %>% # We do not want to handle 's (shouldn't happen because of earlier cleaning, but just in case)
  mutate(word_root = str_to_lower(word_root)) %>% # Converting roots (lemmas) to lowercase (should do automatically, but extra precaution)
  filter(nchar(word_root) > 2) # Removing any words with a lemma of 2 characters or less 
```

### Visualizing output
```{r}
head(pokemon_words, 25) # visualizing
```

## Additional word cleaning
```{r}
# Preparing vectors of unqualified descriptive words
# Pokemon names - extracting from data itself
poke_names <- unique(str_to_lower(str_trim(pokemon_updated$name, side = "both")))

# Pokemon region names (one per game/generation) - googled these
poke_regions <- c("kanto", "johto", "hoenn", "sinnoh", "unova", "kalos", "alola", "galar", "paldea")

# Pokemon types - extracting from data itself
poke_types <- unique(pokemon_updated$type_1)

# General in-game Pokemon terminology (in their lemma form)
poke_game_terms <- c("pokemon", "version", "game", "attack", "battle", "region", "evolve", "move", "mega", "legendary")  

pokemon_words <- pokemon_words %>% 
  # Removal Rule 1: Removing any words that are generation specific (region name), that are the names of types, that are the names of pokemon, and that are basic pokemon game mechanic language
  mutate(word_root = str_trim(word_root, side = "both")) %>% # making sure no whitespace messes up the filtering
  filter(!word_root %in% poke_names) %>% # removing pokemon names
  filter(!word_root %in% poke_regions) %>% # removing pokemon game regions
  filter(!word_root %in% poke_types) %>% # removing pokemon types
  filter(!word_root %in% poke_game_terms) # removing pokemon in-game terminology

# Loading filler words from tidytext
data("stop_words") # returns a dataframe of various stop words (AKA filler words) in English which are not informative

pokemon_words <- pokemon_words %>% 
  # Removal Rule 2: Removing any basic filler words using "stop_words"
  anti_join(stop_words, by = c("word_root" = "word")) %>%  # essentially removes the words that have crossover, in this case fillers
  # Rule 3: Removing any word fragment that may exist like hp
  filter(nchar(word_root) > 2) # removing very short word fragments (did this earlier too, but incase they made it through because why not)

# Checking effect of Rule 1
# The following pokemon had examples of Rule 1 cases
# Ekans has pokemon names in it; Dialga has region name in it; Wartortle has type in it
print("Checking Rule 1 using the following examples:\n\tEkans for pokemon names\n\tDialga for region name\n\tWartortle for type name")
print(pokemon_words[pokemon_words$name %in% c("ekans", "dialga", "wartortle"), ])


# Joining the cleaned words with the rest of the data
pokemon_cleaned <- left_join(pokemon_smart_clean, pokemon_words, by = "name") %>% 
  select(-pokedex_entry) # removing the pokedex entry variable that wasn't used when extracting words (keeping the cleaned one we fed into function instead)

# Making missings all be NA (will need this once we pivot-long for type data)
pokemon_cleaned[pokemon_cleaned == ""] <- NA
```

## Applying TF-IDF
**Explaining terminology:**   
TF (Term Frequency) = how often a word appears within a Pokemon type (count of word in type / count of all words in type)

IDF (Inverse Document Frequency) = how rare a word is across all Pokemon types (math is log(n_types/n_types using word at least once)) - there are 18 types (`pokemon_long`)

TF_IDF = how unique a word is for a type (TF * IDF), will be higher when a word appears a lot within a type (high TF) and when it only appears in a select few types (high IDF)


### Type-based profiles computations
```{r}
# Making type long so that dual-types are considered in the counts for both types
pokemon_type_long <- pokemon_cleaned %>% 
  pivot_longer(
    cols = c(type_1, type_2),
    names_to = "type_slot",
    values_to = "type",
    values_drop_na = TRUE # this way pokemon with only one type stick to having one row
  )

# Creating type profiles by using tf-idf
type_profiles <- pokemon_type_long %>% 
  group_by(type, word_root) %>% # want a unique tf-idf for each word for each pokemon type
  summarise(
    n = n(), # will say how many times the word appears in that type
    n_pokemon = n_distinct(name), # will say how many pokemon of that type have this word in their description
    .groups = "drop"
  ) %>% 
  filter(n_pokemon >= 2) %>% # Setting condition that at least two pokemon need to have that descriptive word within their entry
  bind_tf_idf(word_root, type, n) %>% # calculating tf-idf
  arrange(type, desc(tf_idf)) # Organizing how it looks so higher tf-idf appears first

# Calculating average TF-IDF score across types
type_word_scores <- type_profiles %>% 
  group_by(word_root) %>% 
  summarise(
    type_mean_tfidf = mean(tf_idf)
  ) %>% 
  arrange(desc(type_mean_tfidf))

# Checking basic things:
cat("In the type-profiles there are", length(unique(type_profiles$type)), "unique types.") # should indicate 18 unique types

# Getting an idea of how many unique words there are in general
cat("\n\nThere are", length(unique(type_profiles$word_root)), "unique words in the type-profiles")
```

### Generation-based profile computations
```{r}
# Creating generation profiles by using tf-idf

# Check the type-profiles for code explanation (it is identical except for using generation instead of type)
gen_profiles <- pokemon_cleaned %>% 
  group_by(generation, word_root) %>% 
  summarise(
    n = n(),
    n_pokemon = n_distinct(name),
    .groups = "drop"
  ) %>% 
  filter(n_pokemon >= 2) %>% # Setting condition that at least two pokemon need to have that descriptive word within their entry
  bind_tf_idf(word_root, generation, n) %>% 
  arrange(generation, desc(tf_idf))

# Calculating average TF-IDF score across generations
gen_word_scores <- gen_profiles %>% 
  group_by(word_root) %>% 
  summarise(
    gen_mean_tfidf = mean(tf_idf)
  ) %>% 
  arrange(desc(gen_mean_tfidf))

# Checking basic things:
cat("In the generation-profiles there are", length(unique(gen_profiles$generation)), "unique generations") # should indicate 9 unique generations

# Getting an idea of how many unique words there are in general
cat("\n\nThere are", length(unique(gen_profiles$word_root)), "unique words in the generation-profiles")
```

### Assessing TF-IDF
```{r}
print("Top 20 TF-IDF words in type-based profiles")
print(head(type_word_scores, 20))

print("Top 20 TF-IDF words in generation-based profiles")
print(head(gen_word_scores, 20))

# Merging the two
common_words <- type_word_scores %>% 
  inner_join(gen_word_scores, by = "word_root") %>% 
  mutate(
    avg_tfidf = (type_mean_tfidf + gen_mean_tfidf)/2
  ) %>% 
  arrange(desc(avg_tfidf))

print("Top 20 averaged TF-IDF words across type and generation profiles")
print(head(common_words, 20))

# Extracting the top 150 words to use for clustering features
top_150_df <- common_words %>% 
  slice_head(n = 150)

top_150_words <- top_150_df$word_root
```

## Creating matrices
### Type-word matrix
```{r}
# Creating the matrix by pivoting original profiles wide after filtering to only the selected 150 words (we want each row to be a unique type so that we can cluster across types during HCA with the words as features)
type_matrix <- type_profiles %>% 
  filter(word_root %in% top_150_words) %>% 
  select(type, word_root, tf_idf) %>% 
  rename(pokemon_type = type) %>% 
  pivot_wider(names_from = word_root,
              values_from = tf_idf,
              values_fill = 0) # filling in cases where a word is not reported in a type to be 0

print(type_matrix)

# Preparing quick visualization of the number of types that have a word value greater than 0 to get an idea of how exclusive words are to certain types
word_type_counts <- type_matrix %>% 
  summarise(across(-pokemon_type, 
                   ~ sum(. > 0))) %>% 
  pivot_longer(
    cols = everything(),
    names_to = "word_root",
    values_to =  "types_with_word"
  )

print(word_type_counts)
```

### Generation-word matrix
```{r}
# Creating the matrix by pivoting original profiles wide after filtering to only the selected 150 words (we want each row to be a unique generation so that we can cluster across types during HCA with the words as features)
gen_matrix <- gen_profiles %>% 
  filter(word_root %in% top_150_words) %>% 
  select(generation, word_root, tf_idf) %>% 
  rename(pokemon_gen = generation) %>% 
  pivot_wider(names_from = word_root,
              values_from = tf_idf,
              values_fill = 0) # filling in cases where a word is not reported in a type to be 0

print(gen_matrix)

# Preparing quick visualization of the number of generations that have a word value greater than 0 to get an idea of how exclusive words are to certain generations
word_gen_counts <- gen_matrix %>% 
  summarise(across(-pokemon_gen, 
                   ~ sum(. > 0))) %>% 
  pivot_longer(
    cols = everything(),
    names_to = "word_root",
    values_to =  "gens_with_word"
  )

print(word_gen_counts)
```
### Extra checks and descriptives
```{r}
# checking that the words used in the two matrices are the same for comparison purposes
cat(sum(sort(colnames(type_matrix)[2:151]) == sort(colnames(gen_matrix)[2:151])), "out of 150 words match in the matrices for type and generation.\n\n")

# checking descriptives of the number of types/generations that have a word value greater than 0 to get an idea of how exclusive words are to certain types/generations

cat("Descriptives for number of types with a descriptive word\n")
print(psych::describe(word_type_counts$types_with_word))

cat("\nDescriptives for number of generations with a descriptive word\n")
print(psych::describe(word_gen_counts$gens_with_word))

# Making sure all of the types/generations have a word in the top 150 with a non-zero TF-IDF score
cat("\n", sum(rowSums(is.na(type_matrix[, 2:151])) > 0), "rows in the type-word matrix contain all 0 values.")

cat("\n\n", sum(rowSums(is.na(gen_matrix[, 2:151])) > 0), "rows in the generation-word matrix contain all 0 values.")
```
## Creating Sensitivy Check Matrices
### Extracting new word counts
```{r}
# Extracting the top 50 words to use for clustering features
top_50_words <- common_words %>% 
  slice_head(n = 50) %>% 
  pull(word_root)

# Extracting the top 100 words to use for clustering features
top_100_words <- common_words %>% 
  slice_head(n = 100) %>% 
  pull(word_root)

# Extracting the top 200 words to use for clustering features
top_200_words <- common_words %>% 
  slice_head(n = 200) %>% 
  pull(word_root)
```

### Type-word matrix
```{r}
# Creating the matrix by pivoting original profiles wide after filtering to only the selected 50/100/200 words (we want each row to be a unique type so that we can cluster across types during HCA with the words as features)

# 50 WORDS #################
type_matrix50 <- type_profiles %>% 
  filter(word_root %in% top_50_words) %>% 
  select(type, word_root, tf_idf) %>% 
  rename(pokemon_type = type) %>% 
  pivot_wider(names_from = word_root,
              values_from = tf_idf,
              values_fill = 0) # filling in cases where a word is not reported in a type to be 0

print(type_matrix50)

# Preparing quick visualization of the number of types that have a word value greater than 0 to get an idea of how exclusive words are to certain types
word_type_counts50 <- type_matrix50 %>% 
  summarise(across(-pokemon_type, 
                   ~ sum(. > 0))) %>% 
  pivot_longer(
    cols = everything(),
    names_to = "word_root_50total",
    values_to =  "types_with_word"
  )

print(word_type_counts50)

# 100 WORDS #################
type_matrix100 <- type_profiles %>% 
  filter(word_root %in% top_100_words) %>% 
  select(type, word_root, tf_idf) %>% 
  rename(pokemon_type = type) %>% 
  pivot_wider(names_from = word_root,
              values_from = tf_idf,
              values_fill = 0) # filling in cases where a word is not reported in a type to be 0

print(type_matrix100)

# Preparing quick visualization of the number of types that have a word value greater than 0 to get an idea of how exclusive words are to certain types
word_type_counts100 <- type_matrix100 %>% 
  summarise(across(-pokemon_type, 
                   ~ sum(. > 0))) %>% 
  pivot_longer(
    cols = everything(),
    names_to = "word_root_100total",
    values_to =  "types_with_word"
  )

print(word_type_counts100)

# 200 WORDS #################
type_matrix200 <- type_profiles %>% 
  filter(word_root %in% top_200_words) %>% 
  select(type, word_root, tf_idf) %>% 
  rename(pokemon_type = type) %>% 
  pivot_wider(names_from = word_root,
              values_from = tf_idf,
              values_fill = 0) # filling in cases where a word is not reported in a type to be 0

print(type_matrix200)

# Preparing quick visualization of the number of types that have a word value greater than 0 to get an idea of how exclusive words are to certain types
word_type_counts200 <- type_matrix200 %>% 
  summarise(across(-pokemon_type, 
                   ~ sum(. > 0))) %>% 
  pivot_longer(
    cols = everything(),
    names_to = "word_root_200total",
    values_to =  "types_with_word"
  )

print(word_type_counts200)
```

### Generation-word matrix
```{r}
# Creating the matrix by pivoting original profiles wide after filtering to only the selected 50/100/200 words (we want each row to be a unique generation so that we can cluster across types during HCA with the words as features)

# 50 WORDS #################
gen_matrix50 <- gen_profiles %>% 
  filter(word_root %in% top_50_words) %>% 
  select(generation, word_root, tf_idf) %>% 
  rename(pokemon_gen = generation) %>% 
  pivot_wider(names_from = word_root,
              values_from = tf_idf,
              values_fill = 0) # filling in cases where a word is not reported in a type to be 0

print(gen_matrix50)

# Preparing quick visualization of the number of generations that have a word value greater than 0 to get an idea of how exclusive words are to certain generations
word_gen_counts50 <- gen_matrix50 %>% 
  summarise(across(-pokemon_gen, 
                   ~ sum(. > 0))) %>% 
  pivot_longer(
    cols = everything(),
    names_to = "word_root_50total",
    values_to =  "gens_with_word"
  )

print(word_gen_counts50)

# 100 WORDS #################
gen_matrix100 <- gen_profiles %>% 
  filter(word_root %in% top_100_words) %>% 
  select(generation, word_root, tf_idf) %>% 
  rename(pokemon_gen = generation) %>% 
  pivot_wider(names_from = word_root,
              values_from = tf_idf,
              values_fill = 0) # filling in cases where a word is not reported in a type to be 0

print(gen_matrix100)

# Preparing quick visualization of the number of generations that have a word value greater than 0 to get an idea of how exclusive words are to certain generations
word_gen_counts100 <- gen_matrix100 %>% 
  summarise(across(-pokemon_gen, 
                   ~ sum(. > 0))) %>% 
  pivot_longer(
    cols = everything(),
    names_to = "word_root_100total",
    values_to =  "gens_with_word"
  )

print(word_gen_counts100)

# 200 WORDS #################
gen_matrix200 <- gen_profiles %>% 
  filter(word_root %in% top_200_words) %>% 
  select(generation, word_root, tf_idf) %>% 
  rename(pokemon_gen = generation) %>% 
  pivot_wider(names_from = word_root,
              values_from = tf_idf,
              values_fill = 0) # filling in cases where a word is not reported in a type to be 0

print(gen_matrix200)

# Preparing quick visualization of the number of generations that have a word value greater than 0 to get an idea of how exclusive words are to certain generations
word_gen_counts200 <- gen_matrix200 %>% 
  summarise(across(-pokemon_gen, 
                   ~ sum(. > 0))) %>% 
  pivot_longer(
    cols = everything(),
    names_to = "word_root_200total",
    values_to =  "gens_with_word"
  )

print(word_gen_counts200)
```

# Saving matrices for analysis
```{r}
# Main analysis (150 words)
write.csv(type_matrix, here("project1/DATA/", "pokemon_type-word_matrix.csv"), row.names = FALSE)
write.csv(gen_matrix, here("project1/DATA/", "pokemon_gen-word_matrix.csv"), row.names = FALSE)

# Sensitivity check
# 50 words
write.csv(type_matrix50, here("project1/DATA/", "pokemon_type-word_matrix50.csv"), row.names = FALSE)
write.csv(gen_matrix50, here("project1/DATA/", "pokemon_gen-word_matrix50.csv"), row.names = FALSE)

# 100 words
write.csv(type_matrix100, here("project1/DATA/", "pokemon_type-word_matrix100.csv"), row.names = FALSE)
write.csv(gen_matrix100, here("project1/DATA/", "pokemon_gen-word_matrix100.csv"), row.names = FALSE)

# 200 words
write.csv(type_matrix200, here("project1/DATA/", "pokemon_type-word_matrix200.csv"), row.names = FALSE)
write.csv(gen_matrix200, here("project1/DATA/", "pokemon_gen-word_matrix200.csv"), row.names = FALSE)
```
